Practical Optional Types for Clojure

### 2015-02-20 formal-model.tex
multimethod

Types include the top type, untagged unions, functions, singleton types
and class instances. We abbreviate \EmptyUnion{} as \Bot{}.
Function types contain \emph{latent} propositions and object, which, along with the return type,
may refer to the parameter identifier. They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

@@

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{figure:termsyntax}
\end{figure*}

@@

To sketch how this example is checked, first T-Abs type checks the body under the 
environment \propenv{} = {\isprop{\Union{\nil}{\Number{}}}{\x{}}}.
T-Do then extracts the propositions of its first argument:
T-If tells us \thenprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} and 
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
using T-Error, T-Nil and T-Subsume.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \mintinline{clojure}{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}
and by L-Update we know {\isprop{\Number}{\x{}}}
so T-Local says \mintinline{clojure}{x} is a \Number.

@@

The reason to peel off one path element is subtle. For example,
claiming {\isprop{\Value{\Object}} {\path{\classpe{}}{\x{}}}}
might seem suitable, however it implies \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Object}
which is wrong if \x{} = \makekw{a} where \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Keyword}.
(In formulating \isacompareliteral we found a three year old bug in how core.typed handles this case).

@@

\begin{figure}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt \multimethodexpsyntax{} 
                &\mbox{Expressions} \\
  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
                &\mbox{Values} \\

 \disptablesyntax{} \\
\end{altgrammar}
$$
\caption{Multimethod Syntax}
\end{figure}

@@

  \TDefMethod{}
\end{mathpar}
\caption{Multimethod typing rules}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \getmethodfigure{}
\begin{mathpar}

       \BDefMulti{}

@@

  \BBetaMulti{}
\end{mathpar}
\caption{Multimethod semantics}
\end{figure}

Clojure provides multimethods which can dispatch on the result of an
arbitrary function, the \emph{dispatch function}. \emph{Methods} are
added to the multimethod
which are associated with a \emph{dispatch value}, which is compared to the
result of apply the dispatch function to the argument passed to the
multimethod. If this comparison unambiguously dispatches to one method,
that method is then called with the current arguments.

Occurrence typing, in particular when extended with paths,
for checking method definitions for arbitrary dispatch functions.

\begin{lstlisting}
(ns example.mm
  (:require 
    [clojure.core.typed :refer [ann]]))

(ann my-mm [Any -> Number])
(defmulti my-mm class)

(defmethod my-mm Float [f] (inc f))
\end{lstlisting}

\begin{lstlisting}
(ann clojure.core/class
  [Any -> (U nil Class)
   :object {:id 0, :path [Class]}])
\end{lstlisting}

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}\\
            &\alt&  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
    &\mbox{Values} \\
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{figure:reflectivesyntax}
\end{figure}

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}\\
    &&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
    \ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \

    \mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Non-Reflective Java Interoperability Syntax}
  \label{figure:nonreflectivesyntax}
\end{figure}

>>>>

Types include the top type, untagged unions, functions, singleton types
and class instances. We abbreviate \EmptyUnion{} as \Bot{}, \Value{\Nil{}} as \Nil{} and similarly
for \True and \False.
Function types contain \emph{latent} propositions and object, which, along with the return type,
may refer to the parameter identifier. They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

@@

                &\mbox{Constants}\\
  \s{}, \t{}    &::=& \Top 
                      \alt {\Unionsplice {\overrightarrow{\t{}}}}
                      \alt
                      {\ArrowOne {\x{}} {\t{}}
                                   {\t{}}
                                   {\filterset {\prop{}} {\prop{}}}
                                   {\object{}}}
                      \alt {\Value \singletonmeta{}} 
                      \alt \trdiff{\class{}}
                &\mbox{Types} \\
  \singletonallsyntax{}
                \\ \\
  \occurrencetypingsyntax{}\\
  \openvsyntax{}
  %\\
  %\classliteralallsyntax{}
\end{altgrammar}
$$
\caption{Syntax of Terms, Types, Propositions and Objects}
\label{figure:termsyntax}
\end{figure*}

@@

To sketch how this example is checked, first T-Abs type checks the body under the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
T-Do then extracts the propositions of its first argument:
T-If tells us \thenprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} and 
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
using T-Error, T-Nil and T-Subsume.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \mintinline{clojure}{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}
and by L-Update we know {\isprop{\Number}{\x{}}}
so T-Local says \mintinline{clojure}{x} is a \Number.

@@

The reason to peel off one path element is subtle. For example,
claiming {\isprop{\Value{\Object}} {\path{\classpe{}}{\x{}}}}
might seem suitable, however it implies \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Object}
which is wrong if \x{} = \makekw{a} where \inopenv{\openv{}}{\path{\classpe{}}{\x{}}}{\Keyword}.
(Formulating \isacompareliteral revealed a three year old bug in the implementation).

@@

\begin{figure*}
  \footnotesize
$$
\begin{altgrammar}
  \e{} &::=& \ldots \alt \multimethodexpsyntax{} 
                &\mbox{Expressions} \\
  \v{} &::=& \ldots \alt {\multi {\v{}} {\disptable{}}}
                &\mbox{Values} \\
  \s{}, \t{} &::=& \ldots \alt {\MultiFntype{\t{}}{\t{}}}
                &\mbox{Types} \\

 \disptablesyntax{} \\
\end{altgrammar}
$$
\caption{Multimethod Syntax}
\label{figure:mmsyntax}
\end{figure*}

@@

    \TDefMethod{}
  \end{mathpar}
  \caption{Multimethod typing rules}
  \label{figure:mmtyping}
\end{figure}

@@

\begin{figure}
  \footnotesize
  \getmethodfigure{}
  \begin{mathpar}
    \BDefMulti{}

@@

    \BBetaMulti{}
  \end{mathpar}
  \caption{Multimethod semantics}
  \label{figure:mmsemantics}
\end{figure}

A multimethod in Clojure is a function with a dispatch mechanism to
of internal methods. Multimethods are created with {\mintinline{clojure}{defmulti}}
and methods are installed via {\mintinline{clojure}{defmethod}}.
They are called like regular functions.

\begin{minted}{clojure}
(ann foo [Any -> String])
(defmulti foo class)
(defmethod foo Number [x] (str (inc x)))
(defmethod foo Keyword [x] (str (name x)))
(foo :a) ;=> "a"
(foo 1)  ;=> "2"
\end{minted}

To ease presentation, we present \emph{immutable}
multimethods (figure~\ref{figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. The above is now written

%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\mintinline{clojure}{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
%                       {\classconst{}}}
%                     {\\ bar}$

\begin{minted}{clojure}
(let [foo (defmulti [Any -> String] class)
      foo (defmethod foo Number [x] (str (inc x)))
      foo (defmethod foo Keyword [x] (str (name x)))]
  (foo :a)) ;=> "a"
\end{minted}

Multimethod semantics are in figure~\ref{figure:mmsemantics}.
B-DefMulti creates a multimethod with a dispatch function and an empty dispatch table.
B-DefMethod returns a new multimethod with an extended dispatch table.
B-BetaMulti invokes the dispatch function with the evaluated argument to get the dispatch value,
which \getmethodliteral{} uses to choose the method to invoke.

The typing rules (figure~\ref{figure:mmtyping}) utilse the {\MultiFntype{\s{}}{\t{}}}
type, where \s{} is the interface type for beta reduction, and \t{} is the type for
the dispatch function. T-DefMulti uses the provided interface type and dispatch function
to infer the multimethod type. T-DefMethod uses \isacompareliteral{} to infer the proposition
that must be true if this method is being invoked. The example can be type checked without
changes since we infer \isprop{\Number}{\x{}} and \isprop{\Keyword}{\x{}} for the respective
method bodies.

@@

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \reflectiveexpsyntax{}
            \alt  \localhintsyntax{} \alt \lethintsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\classvalue{\classhint{}} {\overrightarrow {\classfieldpair{\fld{}} {\v{}}}}}
    &\mbox{Values} \\
    \tatypesyntax{}\\
    \typehintenvsyntax{}\\
    \classtableallsyntax{}
  \end{altgrammar}
  $$
  \caption{Java Interoperability Terms and Types}
  \label{figure:reflectivesyntax}
\end{figure*}

@@

\begin{figure*}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \mininonreflectiveexpsyntax{}
    %&&\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \
    %\ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \

    &\mbox{Expressions}
  \end{altgrammar}
  $$
  \caption{Internal Language Non-Reflective Java Interoperability Syntax}
  \label{figure:nonreflectivesyntax}
\end{figure*}
